---
title: "SEV Met Yearly nlme Models"
author: "KM Hall"
date: "2022-08-30"
output: html_document
---

This report runs base, AR1, and AR2 models (variable ~ time) for annual SEV meteorological data using the nlme R package. Mean annual air temperature, minimum annual air temperature, maximum annual air temperature, and total annual precipitation are modeled. More sophisticated models will likely be required after this preliminary analysis.  

----  


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(nlme)
library(MuMIn)
library(gridExtra)
library(kableExtra)

# Load yearly data ---------------------------------------------

path_to_files <- "../data/processed_data/"

# load and only keep vars of interest
y <- read_csv(paste0(path_to_files, "met_yearly_gap_filled.csv")) %>% 
  mutate(sta = as.factor(sta)) %>% 
  select(sta:ppt)
```


```{r}
str(y)
```

```{r}
summary(y)
```


----  

## Prepare data for modeling -  

Add a time variable for nlme modeling and split data into individual met stations.  

```{r}

# function to select station and variable and add 'time' var
prepare_data_for_nlme <- function(data, sta) {
  # data = dataframe to use
  # sta = met station (in quotes)
  data %>% 
    filter(sta == {{ sta }}) %>% 
    mutate(time = as.numeric(as.factor(year)))
}


m40 <- prepare_data_for_nlme(y, "40")
m42 <- prepare_data_for_nlme(y, "42")
m49 <- prepare_data_for_nlme(y, "49")
m50 <- prepare_data_for_nlme(y, "50")
```

----  

## Set up functions -

This code produces several functions that will be useful for producing cleaner code.  

```{r}
# function for plotting preliminary yearly graph -----------------------------------------------------------

plot_yearly_prelim <- function(data, var, title, y_axis_label) {
  # data = dataset
  # var = variable (not in quotes)
  # title = title for graph (in quotes)
  # y_axis_label = label for y-axis (in quotes)
  ggplot(data, aes(x = year, y = {{ var }}, col = sta)) +
    geom_line(color = "burlywood") +
    geom_smooth(method = "lm", color = "black", size = 0.6) +
    facet_wrap(~ sta) +
    labs(title = title,
         x = "Year",
         y = y_axis_label)
}

# function for plotting graphs with best nlme model results ------------------------------------------------

plot_yearly_results <- function(data, var, coeffs, title, y_axis_label) {
  # data = dataset
  # var = variable (not in quotes)
  # coeffs = object containing coefficients and p-values of best model
  # title = title for graph (in quotes)
  # y_axis_label = label for y-axis (in quotes)
  ggplot(data, aes(x = year, y = {{ var }})) +
    geom_line(color = "burlywood") +
    geom_smooth(method = "lm", color = "black", size = 0.6) +
    labs(title = title,
         x = "Year",
         y = y_axis_label,
         caption = paste("slope = ", round(coeffs[2,1], 3), "\n(p-value = ", coeffs[2,2], ")"))
}


# functions to run nlme models ----------------------------------------------------------

# base model function
base_nlme_model <- function(data, var) {
  # data = dataframe
  # var = var to model
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      method = "ML",
      na.action = na.omit)
}



# AR1 model function
AR1_nlme_model <- function(data, var) {
  # data = dataframe
  # var = var to model
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      correlation = corARMA(form = ~ time, p = 1),
      method = "ML",
      na.action = na.omit)
}


# AR2 model function
AR2_nlme_model <- function(data, var) {
  # data = dataframe
  # var = var to model
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      correlation = corARMA(form = ~ time, p = 2),
      method = "ML",
      na.action = na.omit)
}
```


----  

# Mean Annual Air Temperature (C) - 

```{r}
plot_yearly_prelim(y, airt, "Annual Mean Air Temperature", "Temperature (C)")
```


### Met 40:  

```{r}
# Met 40:

m40_mbase <- base_nlme_model(m40, "airt")
m40_mAR1 <- AR1_nlme_model(m40, "airt")
m40_mAR2 <- AR2_nlme_model(m40, "airt")
  
```

AICc scores:
```{r}
AICc(m40_mbase, m40_mAR1, m40_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m40_mbase)

m40_coeffs <- summary(m40_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_plot <- plot_yearly_results(m40, airt, m40_coeffs, "Met 40 - Deep Well \nAnnual Mean Air Temperature", "Temperature (C)"))
```


### Met 42:  

```{r}
# Met 42:

m42_mbase <- base_nlme_model(m42, "airt")
m42_mAR1 <- AR1_nlme_model(m42, "airt")
m42_mAR2 <- AR2_nlme_model(m42, "airt")
  
```

AICc scores:
```{r}
AICc(m42_mbase, m42_mAR1, m42_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m42_mbase)

m42_coeffs <- summary(m42_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_plot <- plot_yearly_results(m42, airt, m42_coeffs, "Met 42 - Cerro Montoso \nAnnual Mean Air Temperature", "Temperature (C)"))
```


### Met 49:  

```{r}
# Met 49:

m49_mbase <- base_nlme_model(m49, "airt")
m49_mAR1 <- AR1_nlme_model(m49, "airt")
m49_mAR2 <- AR2_nlme_model(m49, "airt")
  
```

AICc scores:
```{r}
AICc(m49_mbase, m49_mAR1, m49_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m49_mbase)

m49_coeffs <- summary(m49_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_plot <- plot_yearly_results(m49, airt, m49_coeffs, "Met 49 - Five Points \nAnnual Mean Air Temperature", "Temperature (C)"))
```


### Met 50:  

```{r}
# Met 50:

m50_mbase <- base_nlme_model(m50, "airt")
m50_mAR1 <- AR1_nlme_model(m50, "airt")
m50_mAR2 <- AR2_nlme_model(m50, "airt")
  
```

AICc scores:
```{r}
AICc(m50_mbase, m50_mAR1, m50_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m50_mbase)

m50_coeffs <- summary(m50_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_plot <- plot_yearly_results(m50, airt, m50_coeffs, "Met 50 - Blue Grama \nAnnual Mean Air Temperature", "Temperature (C)"))
```



### Summary Plots of Mean Annual Air Temperature

```{r}
# airt results 
grid.arrange(m40_plot, m42_plot, m49_plot, m50_plot, ncol=2)
```



----  

# Minimum Annual Air Temperature (C) - 

```{r}
plot_yearly_prelim(y, minair, "Annual Minimum Air Temperature", "Temperature (C)")
```


### Met 40:  

```{r}
# Met 40:

m40_mbase <- base_nlme_model(m40, "minair")
m40_mAR1 <- AR1_nlme_model(m40, "minair")
m40_mAR2 <- AR2_nlme_model(m40, "minair")
  
```

AICc scores:
```{r}
AICc(m40_mbase, m40_mAR1, m40_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m40_mbase)

m40_coeffs <- summary(m40_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_plot <- plot_yearly_results(m40, minair, m40_coeffs, "Met 40 - Deep Well \nAnnual Minimum Air Temperature", "Temperature (C)"))
```


### Met 42:  

```{r}
# Met 42:

m42_mbase <- base_nlme_model(m42, "minair")
m42_mAR1 <- AR1_nlme_model(m42, "minair")
m42_mAR2 <- AR2_nlme_model(m42, "minair")
  
```

AICc scores:
```{r}
AICc(m42_mbase, m42_mAR1, m42_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m42_mbase)

m42_coeffs <- summary(m42_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_plot <- plot_yearly_results(m42, minair, m42_coeffs, "Met 42 - Cerro Montoso \nAnnual Minimum Air Temperature", "Temperature (C)"))
```


### Met 49:  

```{r}
# Met 49:

m49_mbase <- base_nlme_model(m49, "minair")
m49_mAR1 <- AR1_nlme_model(m49, "minair")
m49_mAR2 <- AR2_nlme_model(m49, "minair")
  
```

AICc scores:
```{r}
AICc(m49_mbase, m49_mAR1, m49_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m49_mbase)

m49_coeffs <- summary(m49_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_plot <- plot_yearly_results(m49, minair, m49_coeffs, "Met 49 - Five Points \nAnnual Minimum Air Temperature", "Temperature (C)"))
```


### Met 50:  

```{r}
# Met 50:

m50_mbase <- base_nlme_model(m50, "minair")
m50_mAR1 <- AR1_nlme_model(m50, "minair")
m50_mAR2 <- AR2_nlme_model(m50, "minair")
  
```

AICc scores:
```{r}
AICc(m50_mbase, m50_mAR1, m50_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m50_mbase)

m50_coeffs <- summary(m50_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_plot <- plot_yearly_results(m50, minair, m50_coeffs, "Met 50 - Blue Grama \nAnnual Minimum Air Temperature", "Temperature (C)"))
```



### Summary Plots of Minimum Annual Air Temperature

```{r}
# minair results 
grid.arrange(m40_plot, m42_plot, m49_plot, m50_plot, ncol=2)
```



----  

# Maximum Annual Air Temperature (C) - 

```{r}
plot_yearly_prelim(y, maxair, "Annual Maximum Air Temperature", "Temperature (C)")
```


### Met 40:  

```{r}
# Met 40:

m40_mbase <- base_nlme_model(m40, "maxair")
m40_mAR1 <- AR1_nlme_model(m40, "maxair")
m40_mAR2 <- AR2_nlme_model(m40, "maxair")
  
```

AICc scores:
```{r}
AICc(m40_mbase, m40_mAR1, m40_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

The AR1 model has the lowest AICc score. This model is used for plotting.    

```{r}
summary(m40_mAR1)

m40_coeffs <- summary(m40_mAR1)$tTable[,c(1, 4)]
```


```{r}
(m40_plot <- plot_yearly_results(m40, maxair, m40_coeffs, "Met 40 - Deep Well \nAnnual Maximum Air Temperature", "Temperature (C)"))
```


### Met 42:  

```{r}
# Met 42:

m42_mbase <- base_nlme_model(m42, "maxair")
m42_mAR1 <- AR1_nlme_model(m42, "maxair")
m42_mAR2 <- AR2_nlme_model(m42, "maxair")
  
```

AICc scores:
```{r}
AICc(m42_mbase, m42_mAR1, m42_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

The AR1 model has the lowest AICc score. This model is used for plotting.  



```{r}
summary(m42_mAR1)

m42_coeffs <- summary(m42_mAR1)$tTable[,c(1, 4)]
```


```{r}
(m42_plot <- plot_yearly_results(m42, maxair, m42_coeffs, "Met 42 - Cerro Montoso \nAnnual Maximum Air Temperature", "Temperature (C)"))
```


### Met 49:  

```{r}
# Met 49:

m49_mbase <- base_nlme_model(m49, "maxair")
m49_mAR1 <- AR1_nlme_model(m49, "maxair")
m49_mAR2 <- AR2_nlme_model(m49, "maxair")
  
```

AICc scores:
```{r}
AICc(m49_mbase, m49_mAR1, m49_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

The base model has the lowest AICc and is used for plotting.  

```{r}
summary(m49_mbase)

m49_coeffs <- summary(m49_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_plot <- plot_yearly_results(m49, maxair, m49_coeffs, "Met 49 - Five Points \nAnnual Maximum Air Temperature", "Temperature (C)"))
```


### Met 50:  

```{r}
# Met 50:

m50_mbase <- base_nlme_model(m50, "maxair")
m50_mAR1 <- AR1_nlme_model(m50, "maxair")
m50_mAR2 <- AR2_nlme_model(m50, "maxair")
  
```

AICc scores:
```{r}
AICc(m50_mbase, m50_mAR1, m50_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

The AR1 model has the lowest AICc score. This model is used for plotting.  


```{r}
summary(m50_mAR1)

m50_coeffs <- summary(m50_mAR1)$tTable[,c(1, 4)]
```


```{r}
(m50_plot <- plot_yearly_results(m50, maxair, m50_coeffs, "Met 50 - Blue Grama \nAnnual Maximum Air Temperature", "Temperature (C)"))
```



### Summary Plots of Maximum Annual Air Temperature

```{r}
# maxair results 
grid.arrange(m40_plot, m42_plot, m49_plot, m50_plot, ncol=2)
```



----  

# Total Annual Precipitation (mm) - 

```{r}
plot_yearly_prelim(y, ppt, "Total Annual Precipitation", "Precipitation (mm)")
```


### Met 40:  

```{r}
# Met 40:

m40_mbase <- base_nlme_model(m40, "ppt")
m40_mAR1 <- AR1_nlme_model(m40, "ppt")
m40_mAR2 <- AR2_nlme_model(m40, "ppt")
  
```

AICc scores:
```{r}
AICc(m40_mbase, m40_mAR1, m40_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m40_mbase)

m40_coeffs <- summary(m40_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_plot <- plot_yearly_results(m40, ppt, m40_coeffs, "Met 40 - Deep Well \nTotal Annual Precipitation", "Precipitation (mm)"))
```


### Met 42:  

```{r}
# Met 42:

m42_mbase <- base_nlme_model(m42, "ppt")
m42_mAR1 <- AR1_nlme_model(m42, "ppt")
m42_mAR2 <- AR2_nlme_model(m42, "ppt")
  
```

AICc scores:
```{r}
AICc(m42_mbase, m42_mAR1, m42_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m42_mbase)

m42_coeffs <- summary(m42_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_plot <- plot_yearly_results(m42, ppt, m42_coeffs, "Met 42 - Cerro Montoso \nTotal Annual Precipitation", "Precipitation (mm)"))
```


### Met 49:  

```{r}
# Met 49:

m49_mbase <- base_nlme_model(m49, "ppt")
m49_mAR1 <- AR1_nlme_model(m49, "ppt")
m49_mAR2 <- AR2_nlme_model(m49, "ppt")
  
```

AICc scores:
```{r}
AICc(m49_mbase, m49_mAR1, m49_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m49_mbase)

m49_coeffs <- summary(m49_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_plot <- plot_yearly_results(m49, ppt, m49_coeffs, "Met 49 - Five Points \nTotal Annual Precipitation", "Precipitation (mm)"))
```


### Met 50:  

```{r}
# Met 50:

m50_mbase <- base_nlme_model(m50, "ppt")
m50_mAR1 <- AR1_nlme_model(m50, "ppt")
m50_mAR2 <- AR2_nlme_model(m50, "ppt")
  
```

AICc scores:
```{r}
AICc(m50_mbase, m50_mAR1, m50_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```


```{r}
summary(m50_mbase)

m50_coeffs <- summary(m50_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_plot <- plot_yearly_results(m50, ppt, m50_coeffs, "Met 50 - Blue Grama \nTotal Annual Precipitation", "Precipitation (mm)"))
```



### Summary Plots of Total Annual Precipitation

```{r}
# ppt results 
grid.arrange(m40_plot, m42_plot, m49_plot, m50_plot, ncol=2)
```

