---
title: "SEV Met Monthly nlme Models"
author: "KM Hall"
date: "2022-08-30"
output: html_document
---

This report runs base, AR1, and AR2 models (variable ~ time) for monthly SEV meteorological data using the nlme R package. Mean monthly air temperature, minimum monthly air temperature, maximum monthly air temperature, and total monthly precipitation are modeled. More sophisticated models will likely be required after this preliminary analysis. The main goal is to determine whether certain variables are changing over time for different months of the year for the SEV meteorological stations.


----  


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(nlme)
library(MuMIn)
library(gridExtra)
library(kableExtra)

# Load yearly data ---------------------------------------------

path_to_files <- "../data/processed_data/"

# load and only keep vars of interest
m <- read_csv(paste0(path_to_files, "met_monthly_gap_filled.csv")) %>% 
  mutate(sta = as.factor(sta)) %>% 
  select(sta:ppt)
```


```{r}
str(m)
```


```{r}
summary(m)
```


----  


## Prepare data for modeling -  

Split data into individual met stations and add a time variable for nlme modeling.  

```{r}

# function to select station and variable and add 'time' var for a specific month of the year
prepare_data_for_nlme <- function(data, sta, month) {
  # data = dataframe to use
  # sta = met station (in quotes)
  # month = month of year (in quotes)
  data %>% 
    filter(sta == {{ sta }} & month == {{ month }}) %>% 
    arrange(date) %>% 
    mutate(time = as.numeric(as.factor(year)))
}


```


----  

## Set up functions -

This code produces several functions that will be useful for producing cleaner code.  

```{r}
# function for plotting preliminary monthly graph -----------------------------------------------------------

plot_monthly_prelim <- function(data, var, title, y_axis_label) {
  # data = dataset
  # var = variable (not in quotes)
  # title = title for graph (in quotes)
  # y_axis_label = label for y-axis (in quotes)
  ggplot(data, aes(x = date, y = {{ var }}, col = sta)) +
    geom_line(color = "burlywood") +
    geom_smooth(method = "lm", color = "black", size = 0.6) +
    facet_wrap(~ sta) +
    labs(title = title,
         x = "Month",
         y = y_axis_label)
}



by_month_lm <- function(data, var, station, title, y_axis_label) {
  # data = dataset
  # var = variable to plot (not in quotes)
  # station = met station (in quotes)
  # title = graph title
  # y_axis_label = label for the graph's y-axis
  m %>% 
    filter(sta == station) %>% 
    ggplot(., aes(x = year, y = {{ var }})) +
    geom_line() +
    geom_smooth(method = "lm") +
    facet_wrap(~ month, scales = "free_y") +
    ggtitle({{ title }})
}


# function for plotting graphs with best nlme model results ------------------------------------------------

plot_monthly_results <- function(data, var, coeffs, title, y_axis_label) {
  # data = dataset
  # var = variable (not in quotes)
  # coeffs = object containing coefficients and p-values of best model
  # title = title for graph (in quotes)
  # y_axis_label = label for y-axis (in quotes)
  ggplot(data, aes(x = date, y = {{ var }})) +
    geom_line(color = "burlywood") +
    geom_smooth(method = "lm", color = "black", size = 0.6) +
    labs(title = title,
         x = "Month",
         y = y_axis_label,
         caption = paste("slope = ", round(coeffs[2,1], 3), "\n(p-value = ", coeffs[2,2], ")"))
}


# functions to run nlme models ----------------------------------------------------------

# base model function
base_nlme_model <- function(data, var, month) {
  # data = dataframe
  # var = var to model
  # month = month of year
  
  data <- data %>% filter(month == month)
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      method = "ML",
      na.action = na.omit)
}



# AR1 model function
AR1_nlme_model <- function(data, var, month) {
  # data = dataframe
  # var = var to model
  # month = month of year
  
  data <- data %>% filter(month == month)
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      correlation = corARMA(form = ~ time, p = 1),
      method = "ML",
      na.action = na.omit)
}


# AR2 model function
AR2_nlme_model <- function(data, var, month) {
  # data = dataframe
  # var = var to model
  # month = month of year
  
  data <- data %>% filter(month == month)
  
  model_specification = as.formula(paste0(var, " ~ time"))
  
  gls(model = model_specification,
      data = data,
      correlation = corARMA(form = ~ time, p = 2),
      method = "ML",
      na.action = na.omit)
}
```


----  

# Mean Monthly Air Temperature -  



```{r}
plot_monthly_prelim(m, airt, "Monthly Mean Air Temperature", "Temperature (C)")
```


### Met 40 - June:  

```{r}
by_month_lm(m, airt, "40", "Met 40 - Deep Well \nMean Monthly Air Temperature", "Temperature (C)")
```


```{r}
m40_6 <- prepare_data_for_nlme(m, "40", "6")

m40_6_mbase <- base_nlme_model(m40_6, "airt", 6)
m40_6_mAR1 <- AR1_nlme_model(m40_6, "airt", 6)
m40_5_mAR2 <- AR2_nlme_model(m40_6, "airt", 6)
```

AICc scores:
```{r}
AICc(m40_6_mbase, m40_6_mAR1, m40_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m40_6_mbase)

m40_6_coeffs <- summary(m40_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_6_plot <- plot_monthly_results(m40_6, airt, m40_6_coeffs, "Met 40 - Deep Well \nJune Mean Air Temperature", "Temperature (C)"))
```


### Met 42 - June:  

```{r}
by_month_lm(m, airt, "42", "Met 42 - Deep Well \nMean Monthly Air Temperature", "Temperature (C)")
```


```{r}
m42_6 <- prepare_data_for_nlme(m, "42", "6")

m42_6_mbase <- base_nlme_model(m42_6, "airt", 6)
m42_6_mAR1 <- AR1_nlme_model(m42_6, "airt", 6)
m42_5_mAR2 <- AR2_nlme_model(m42_6, "airt", 6)
```

AICc scores:
```{r}
AICc(m42_6_mbase, m42_6_mAR1, m42_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m42_6_mbase)

m42_6_coeffs <- summary(m42_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_6_plot <- plot_monthly_results(m42_6, airt, m42_6_coeffs, "Met 42 - Cerro Montoso \nJune Mean Air Temperature", "Temperature (C)"))
```


### Met 49 - June:  

```{r}
by_month_lm(m49_6, airt, "49", "Met 49 - Deep Well \nMean Monthly Air Temperature", "Temperature (C)")
```


```{r}
m49_6 <- prepare_data_for_nlme(m, "49", "6")

m49_6_mbase <- base_nlme_model(m49_6, "airt", 6)
m49_6_mAR1 <- AR1_nlme_model(m49_6, "airt", 6)
m49_5_mAR2 <- AR2_nlme_model(m49_6, "airt", 6)
```

AICc scores:
```{r}
AICc(m49_6_mbase, m49_6_mAR1, m49_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m49_6_mbase)

m49_6_coeffs <- summary(m49_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_6_plot <- plot_monthly_results(m49_6, airt, m49_6_coeffs, "Met 49 - Five Points \nJune Mean Air Temperature", "Temperature (C)"))
```


### Met 50 - June:  

```{r}
by_month_lm(m, airt, "50", "Met 50 - Deep Well \nMean Monthly Air Temperature", "Temperature (C)")
```


```{r}
m50_6 <- prepare_data_for_nlme(m, "50", "6")

m50_6_mbase <- base_nlme_model(m50_6, "airt", 6)
m50_6_mAR1 <- AR1_nlme_model(m50_6, "airt", 6)
m50_5_mAR2 <- AR2_nlme_model(m50_6, "airt", 6)
```

AICc scores:
```{r}
AICc(m50_6_mbase, m50_6_mAR1, m50_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m50_6_mbase)

m50_6_coeffs <- summary(m50_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_6_plot <- plot_monthly_results(m50_6, airt, m50_6_coeffs, "Met 50 - Blue Grama \nJune Mean Air Temperature", "Temperature (C)"))
```


### Summary Plots of Mean Monthly Air Temperature for June

```{r}
# airt results 
(airt_6_graphs <- grid.arrange(m40_6_plot, m42_6_plot, m49_6_plot, m50_6_plot, ncol=2))
```




----  

# Maximum Monthly Air Temperature -  



```{r}
plot_monthly_prelim(m, maxair, "Monthly Maximum Air Temperature", "Temperature (C)")
```


### Met 40 - June:  

```{r}
by_month_lm(m, maxair, "40", "Met 40 - Deep Well \nMaximum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m40_6 <- prepare_data_for_nlme(m, "40", "6")

m40_6_mbase <- base_nlme_model(m40_6, "maxair", 6)
m40_6_mAR1 <- AR1_nlme_model(m40_6, "maxair", 6)
m40_5_mAR2 <- AR2_nlme_model(m40_6, "maxair", 6)
```

AICc scores:
```{r}
AICc(m40_6_mbase, m40_6_mAR1, m40_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m40_6_mbase)

m40_6_coeffs <- summary(m40_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_6_plot <- plot_monthly_results(m40_6, maxair, m40_6_coeffs, "Met 40 - Deep Well \nJune Maximum Air Temperature", "Temperature (C)"))
```


### Met 42 - June:  

```{r}
by_month_lm(m, maxair, "42", "Met 42 - Deep Well \nMaximum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m42_6 <- prepare_data_for_nlme(m, "42", "6")

m42_6_mbase <- base_nlme_model(m42_6, "maxair", 6)
m42_6_mAR1 <- AR1_nlme_model(m42_6, "maxair", 6)
m42_5_mAR2 <- AR2_nlme_model(m42_6, "maxair", 6)
```

AICc scores:
```{r}
AICc(m42_6_mbase, m42_6_mAR1, m42_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m42_6_mbase)

m42_6_coeffs <- summary(m42_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_6_plot <- plot_monthly_results(m42_6, maxair, m42_6_coeffs, "Met 42 - Cerro Montoso \nJune Maximum Air Temperature", "Temperature (C)"))
```


### Met 49 - June:  

```{r}
by_month_lm(m49_6, maxair, "49", "Met 49 - Deep Well \nMaximum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m49_6 <- prepare_data_for_nlme(m, "49", "6")

m49_6_mbase <- base_nlme_model(m49_6, "maxair", 6)
m49_6_mAR1 <- AR1_nlme_model(m49_6, "maxair", 6)
m49_5_mAR2 <- AR2_nlme_model(m49_6, "maxair", 6)
```

AICc scores:
```{r}
AICc(m49_6_mbase, m49_6_mAR1, m49_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m49_6_mbase)

m49_6_coeffs <- summary(m49_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_6_plot <- plot_monthly_results(m49_6, maxair, m49_6_coeffs, "Met 49 - Five Points \nJune Maximum Air Temperature", "Temperature (C)"))
```


### Met 50 - June:  

```{r}
by_month_lm(m, maxair, "50", "Met 50 - Deep Well \nMaximum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m50_6 <- prepare_data_for_nlme(m, "50", "6")

m50_6_mbase <- base_nlme_model(m50_6, "maxair", 6)
m50_6_mAR1 <- AR1_nlme_model(m50_6, "maxair", 6)
m50_5_mAR2 <- AR2_nlme_model(m50_6, "maxair", 6)
```

AICc scores:
```{r}
AICc(m50_6_mbase, m50_6_mAR1, m50_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m50_6_mbase)

m50_6_coeffs <- summary(m50_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_6_plot <- plot_monthly_results(m50_6, maxair, m50_6_coeffs, "Met 50 - Blue Grama \nJune Maximum Air Temperature", "Temperature (C)"))
```


### Summary Plots of Maximum Monthly Air Temperature for June

```{r}
# maxair results 
(maxair_6_graphs <- grid.arrange(m40_6_plot, m42_6_plot, m49_6_plot, m50_6_plot, ncol=2))
```







----  

# Minimum Monthly Air Temperature -  



```{r}
plot_monthly_prelim(m, minair, "Monthly Minimum Air Temperature", "Temperature (C)")
```


### Met 40 - June:  

```{r}
by_month_lm(m, minair, "40", "Met 40 - Deep Well \nMinimum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m40_6 <- prepare_data_for_nlme(m, "40", "6")

m40_6_mbase <- base_nlme_model(m40_6, "minair", 6)
m40_6_mAR1 <- AR1_nlme_model(m40_6, "minair", 6)
m40_5_mAR2 <- AR2_nlme_model(m40_6, "minair", 6)
```

AICc scores:
```{r}
AICc(m40_6_mbase, m40_6_mAR1, m40_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m40_6_mbase)

m40_6_coeffs <- summary(m40_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_6_plot <- plot_monthly_results(m40_6, minair, m40_6_coeffs, "Met 40 - Deep Well \nJune Minimum Air Temperature", "Temperature (C)"))
```


### Met 42 - June:  

```{r}
by_month_lm(m, minair, "42", "Met 42 - Deep Well \nMinimum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m42_6 <- prepare_data_for_nlme(m, "42", "6")

m42_6_mbase <- base_nlme_model(m42_6, "minair", 6)
m42_6_mAR1 <- AR1_nlme_model(m42_6, "minair", 6)
m42_5_mAR2 <- AR2_nlme_model(m42_6, "minair", 6)
```

AICc scores:
```{r}
AICc(m42_6_mbase, m42_6_mAR1, m42_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m42_6_mbase)

m42_6_coeffs <- summary(m42_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_6_plot <- plot_monthly_results(m42_6, minair, m42_6_coeffs, "Met 42 - Cerro Montoso \nJune Minimum Air Temperature", "Temperature (C)"))
```


### Met 49 - June:  

```{r}
by_month_lm(m49_6, minair, "49", "Met 49 - Deep Well \nMinimum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m49_6 <- prepare_data_for_nlme(m, "49", "6")

m49_6_mbase <- base_nlme_model(m49_6, "minair", 6)
m49_6_mAR1 <- AR1_nlme_model(m49_6, "minair", 6)
m49_5_mAR2 <- AR2_nlme_model(m49_6, "minair", 6)
```

AICc scores:
```{r}
AICc(m49_6_mbase, m49_6_mAR1, m49_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m49_6_mAR1)

m49_6_coeffs <- summary(m49_6_mAR1)$tTable[,c(1, 4)]
```


```{r}
(m49_6_plot <- plot_monthly_results(m49_6, minair, m49_6_coeffs, "Met 49 - Five Points \nJune Minimum Air Temperature", "Temperature (C)"))
```


### Met 50 - June:  

```{r}
by_month_lm(m, minair, "50", "Met 50 - Deep Well \nMinimum Monthly Air Temperature", "Temperature (C)")
```


```{r}
m50_6 <- prepare_data_for_nlme(m, "50", "6")

m50_6_mbase <- base_nlme_model(m50_6, "minair", 6)
m50_6_mAR1 <- AR1_nlme_model(m50_6, "minair", 6)
m50_5_mAR2 <- AR2_nlme_model(m50_6, "minair", 6)
```

AICc scores:
```{r}
AICc(m50_6_mbase, m50_6_mAR1, m50_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m50_6_mAR1)

m50_6_coeffs <- summary(m50_6_mAR1)$tTable[,c(1, 4)]
```


```{r}
(m50_6_plot <- plot_monthly_results(m50_6, minair, m50_6_coeffs, "Met 50 - Blue Grama \nJune Minimum Air Temperature", "Temperature (C)"))
```


### Summary Plots of Minimum Monthly Air Temperature for June

```{r}
# minair results 
(minair_6_graphs <- grid.arrange(m40_6_plot, m42_6_plot, m49_6_plot, m50_6_plot, ncol=2))
```



# June Minimum, Mean, and Maximum Air Temperature Results -  

```{r}
minair_6_graphs
```

```{r}
airt_6_graphs
```


```{r}
maxair_6_graphs
```















----  

#  Monthly Precipitation -  



```{r}
plot_monthly_prelim(m, ppt, "Monthly Precipitation", "Precipitation (mm)")
```


### Met 40 - June:  

```{r}
by_month_lm(m, ppt, "40", "Met 40 - Deep Well \n Monthly Precipitation", "Precipitation (mm)")
```


```{r}
m40_6 <- prepare_data_for_nlme(m, "40", "6")

m40_6_mbase <- base_nlme_model(m40_6, "ppt", 6)
m40_6_mAR1 <- AR1_nlme_model(m40_6, "ppt", 6)
m40_5_mAR2 <- AR2_nlme_model(m40_6, "ppt", 6)
```

AICc scores:
```{r}
AICc(m40_6_mbase, m40_6_mAR1, m40_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m40_6_mbase)

m40_6_coeffs <- summary(m40_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m40_6_plot <- plot_monthly_results(m40_6, ppt, m40_6_coeffs, "Met 40 - Deep Well \nJune Precipitation", "Precipitation (mm)"))
```


### Met 42 - June:  

```{r}
by_month_lm(m, ppt, "42", "Met 42 - Deep Well \n Monthly Precipitation", "Precipitation (mm)")
```


```{r}
m42_6 <- prepare_data_for_nlme(m, "42", "6")

m42_6_mbase <- base_nlme_model(m42_6, "ppt", 6)
m42_6_mAR1 <- AR1_nlme_model(m42_6, "ppt", 6)
m42_5_mAR2 <- AR2_nlme_model(m42_6, "ppt", 6)
```

AICc scores:
```{r}
AICc(m42_6_mbase, m42_6_mAR1, m42_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m42_6_mbase)

m42_6_coeffs <- summary(m42_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m42_6_plot <- plot_monthly_results(m42_6, ppt, m42_6_coeffs, "Met 42 - Cerro Montoso \nJune Precipitation", "Precipitation (mm)"))
```


### Met 49 - June:  

```{r}
by_month_lm(m49_6, ppt, "49", "Met 49 - Deep Well \n Monthly Precipitation", "Precipitation (mm)")
```


```{r}
m49_6 <- prepare_data_for_nlme(m, "49", "6")

m49_6_mbase <- base_nlme_model(m49_6, "ppt", 6)
m49_6_mAR1 <- AR1_nlme_model(m49_6, "ppt", 6)
m49_5_mAR2 <- AR2_nlme_model(m49_6, "ppt", 6)
```

AICc scores:
```{r}
AICc(m49_6_mbase, m49_6_mAR1, m49_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m49_6_mbase)

m49_6_coeffs <- summary(m49_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m49_6_plot <- plot_monthly_results(m49_6, ppt, m49_6_coeffs, "Met 49 - Five Points \nJune Precipitation", "Precipitation (mm)"))
```


### Met 50 - June:  

```{r}
by_month_lm(m, ppt, "50", "Met 50 - Deep Well \n Monthly Precipitation", "Precipitation (mm)")
```


```{r}
m50_6 <- prepare_data_for_nlme(m, "50", "6")

m50_6_mbase <- base_nlme_model(m50_6, "ppt", 6)
m50_6_mAR1 <- AR1_nlme_model(m50_6, "ppt", 6)
m50_5_mAR2 <- AR2_nlme_model(m50_6, "ppt", 6)
```

AICc scores:
```{r}
AICc(m50_6_mbase, m50_6_mAR1, m50_5_mAR2) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
summary(m50_6_mbase)

m50_6_coeffs <- summary(m50_6_mbase)$tTable[,c(1, 4)]
```


```{r}
(m50_6_plot <- plot_monthly_results(m50_6, ppt, m50_6_coeffs, "Met 50 - Blue Grama \nJune Precipitation", "Precipitation (mm)"))
```


### Summary Plots of  Monthly Precipitation for June

```{r}
# ppt results 
(ppt_6_graphs <- grid.arrange(m40_6_plot, m42_6_plot, m49_6_plot, m50_6_plot, ncol=2))
```




