---
title: "Test Hourly Gap-Filling"
author: "KM Hall"
date: "6/5/2022"
output: html_document
---


The purpose of this R Markdown notebook is to examine the flags created for the met_hourly_gap_filled.csv file produced in the anomalize_hourly_met_data.R program. There were two types of flags: 1. flags when a variable's hourly data was missing in the raw data, but the hourly record is present in the data; and 2. there was no hourly record in the raw data. 

In the second case, missing hourly records were produced and a no_record_flag was set to TRUE. Except for precipitation (ppt), data was then produced for the record using a simple gap-filling method that looked at a station's mean for each variable on the given julian day and hour of the missing data. Missing ppt data was gap-filled with a value of 0.



----  

```{r, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)
library(kableExtra)

met <- read_csv("../data/processed_data/met_hourly_gap_filled.csv") %>% 
  mutate(sta = as.factor(sta))

```

```{r}
rec_ct <- met %>% 
  group_by(sta) %>% 
  summarize(n_records = n()) %>% 
  ungroup()
```

----  

A no_record_flag of TRUE means that there was not a record in the raw data. There will not be any NAs for the no_record_flag.  

```{r, message=FALSE}
met %>% 
  group_by(sta, no_record_flag) %>% 
  summarize(flag_count = n()) %>% 
  left_join(rec_ct) %>% 
  mutate(flag_percent = flag_count / n_records * 100) %>% 
  select(-n_records) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```





----  

Flags showing how many records are missing for a particular variable in the raw data. TRUE means the data is missing in the raw data, but there is a record present. NA means that the record was not present in the raw data.  

```{r, message=FALSE}
met %>% 
  pivot_longer(airt_miss:maxsol_miss, names_to = "flag", values_to = "flag_value") %>% 
  group_by(sta, flag, flag_value) %>% 
  summarize(flag_count = n()) %>% 
  left_join(rec_ct) %>% 
  mutate(flag_percent = flag_count / n_records * 100) %>% 
  select(-n_records) %>% 
  kbl() %>% 
  kable_styling(full_width = FALSE, position = "left")
```

----   

Test the annual differences in select met variables when not gap-filled versus when gap-filled.  

```{r}

met_flag_diffs <- met %>% 
  mutate(year = year(dt)) %>% 
  group_by(sta, year) %>% 
  summarize(airt_ngf = mean(airt[airt_miss == FALSE], na.rm = TRUE),
            airt_gf = mean(airt),
            airt_diff = airt_ngf - airt_gf,
            minair_ngf = mean(minair[minair_miss == FALSE], na.rm = TRUE),
            minair_gf = mean(minair),
            minair_diff = minair_ngf - minair_gf,
            maxair_ngf = mean(maxair[maxair_miss == FALSE], na.rm = TRUE),
            maxair_gf = mean(maxair),
            maxair_diff = maxair_ngf - maxair_gf,
            rh_ngf = mean(rh[rh_miss == FALSE], na.rm = TRUE),
            rh_gf = mean(rh),
            rh_diff = rh_ngf - rh_gf,
            ppt_ngf = sum(ppt[ppt_miss == FALSE], na.rm = TRUE),
            ppt_gf = sum(ppt),
            ppt_diff = ppt_ngf - ppt_gf,)


```  

----  

airt gap-fill testing:  

```{r}
ggplot(met_flag_diffs, aes(x = year, y = airt_diff, color=sta)) +
  geom_line() +
  facet_wrap(~sta) +
  ggtitle("Mean hourly airt difference between non gap-filled\nand gap-filled data")


met_flag_diffs %>% 
  select(sta, year, airt_ngf, airt_gf) %>% 
  pivot_longer(airt_ngf:airt_gf, names_to = 'gap_fill_status', values_to = 'value') %>% 
  ggplot(aes(x = year, y = value, color = gap_fill_status)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ sta) +
  labs(title = "airt gap-filling")

```


----  

minair gap-fill testing:  

```{r}
ggplot(met_flag_diffs, aes(x = year, y = minair_diff, color=sta)) +
  geom_line() +
  facet_wrap(~sta) +
  ggtitle("Mean hourly minair difference between non gap-filled\nand gap-filled data")

met_flag_diffs %>% 
  select(sta, year, minair_ngf, minair_gf) %>% 
  pivot_longer(minair_ngf:minair_gf, names_to = 'gap_fill_status', values_to = 'value') %>% 
  ggplot(aes(x = year, y = value, color = gap_fill_status)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ sta) +
  labs(title = "minair gap-filling")

```

The low values seen at the beginning of the time-series for station 42 is because minair was not calculated in the raw data. This artificially makes the gap-filling look bad. minair could be approximately calculated from hourly airt data for the early time period where it is missing in the raw data.  

----  
maxair gap-fill testing:  

```{r}
ggplot(met_flag_diffs, aes(x = year, y = maxair_diff, color=sta)) +
  geom_line() +
  facet_wrap(~sta) +
  ggtitle("Mean hourly maxair difference between non gap-filled\nand gap-filled data")

met_flag_diffs %>% 
  select(sta, year, maxair_ngf, maxair_gf) %>% 
  pivot_longer(maxair_ngf:maxair_gf, names_to = 'gap_fill_status', values_to = 'value') %>% 
  ggplot(aes(x = year, y = value, color = gap_fill_status)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ sta) +
  labs(title = "maxair gap-filling")

```

The low values seen at the beginning of the time-series for station 42 is because maxair was not calculated in the raw data. This artificially makes the gap-filling look bad. maxair could be approximately calculated from hourly airt data for the early time period where it is missing in the raw data.  

----  

rh gap-fill testing:  

```{r}
ggplot(met_flag_diffs, aes(x = year, y = rh_diff, color=sta)) +
  geom_line() +
  facet_wrap(~sta) +
  ggtitle("Mean hourly relative humidity difference between non gap-filled\nand gap-filled data")

met_flag_diffs %>% 
  select(sta, year, rh_ngf, rh_gf) %>% 
  pivot_longer(rh_ngf:rh_gf, names_to = 'gap_fill_status', values_to = 'value') %>% 
  ggplot(aes(x = year, y = value, color = gap_fill_status)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ sta) +
  labs(title = "rh gap-filling")

```

----  

ppt gap-fill testing:  

```{r}
ggplot(met_flag_diffs, aes(x = year, y = ppt_diff, color=sta)) +
  geom_line() +
  facet_wrap(~sta) +
  ggtitle("Mean hourly ppt difference between non gap-filled\nand gap-filled data")

met_flag_diffs %>% 
  select(sta, year, ppt_ngf, ppt_gf) %>% 
  pivot_longer(ppt_ngf:ppt_gf, names_to = 'gap_fill_status', values_to = 'value') %>% 
  ggplot(aes(x = year, y = value, color = gap_fill_status)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ sta) +
  labs(title = "ppt gap-filling")

```

Gap-filling precipitation with a value of 0 does not change  total precipitation when gap-filled.
